Decomposing biodiversity change to processes of extinction, colonization, and persistence across scales

### 2023-05-10 scripts/ms_scaling_changes.qmd
initial commits

```{r, message=FALSE, warning=FALSE}
rm(list = ls())
library(caret)
library(caretEnsemble)
library(dplyr)
library(ranger)
library(ggtext)
library(cowplot)
library(forcats)
library(tidyr)
library(stringr)
```

@@

## We perform repeated (n = 3) cross-validation (k = 10) using the caret package
# First we split the dataset into training and validation datasets
# set.seed(333)
traindata <- trainControl(method = "repeatedcv", number = 10, repeats = 3)

@@

## And make predictions for each metric
predicted_nCEP_atlas <- data.frame()
## predictions for each metric
for(THE_metric in names(nCEP_atlas_models)){
  model <- nCEP_atlas_models[[THE_metric]][["xgboost"]]
  pred <- predict(model, 
                  newdata = nCEP_topredict)
  predicted_nCEP_atlas <- 
  rbind(predicted_nCEP_atlas,
        nCEP_topredict %>% cbind(pred = round(pred)) %>% mutate(metric = THE_metric))
}
## Also, compute jaccard index
## Compute turnover
predicted_nCEP_atlas <- 
predicted_nCEP_atlas %>% 
  pivot_wider(id_cols = c(Lat, Long, AREA, polypoint_ratio, KVADRAT, elevation, spatial_scale, effort_diff, timespan_diff),
              names_from = c(metric,period),
              values_from = pred,
              names_sep = "") %>% 
  group_by(Lat, Long, AREA, polypoint_ratio, KVADRAT, elevation, spatial_scale, effort_diff, timespan_diff) %>% 
  mutate(jaccardM2M3 = persM2M3/(persM2M3 + colM2M3 + extM2M3),
         jaccardM3M4 = persM3M4/(persM3M4 + colM3M4 + extM3M4),
         jaccardM2M4 = persM2M4/(persM2M4 + colM2M4 + extM2M4)) %>% 
  ungroup()
```

```{r}
nCEP_BBS_models <- readRDS("../models/nCEP_JPSP.rds")
```

```{r}
predicted_nCEP_BBS <- data.frame()
for(THE_metric in names(nCEP_BBS_models)){
  model <- nCEP_BBS_models[[THE_metric]][["rf"]]
  pred <- 
    predict(model, 
            newdata = JPSP_topredict %>% 
              dplyr::select(-time_span, -id, -sr, -Date, polypoint_ratio, Lat, Long, elevation, AREA) %>% 
              mutate(timespan_diff = 0) %>% 
              crossing(period = c("M2M3", "M3M4", "M2M4")))
  predicted_nCEP_BBS <- 
  rbind(
    predicted_nCEP_BBS,
    JPSP_topredict %>% 
    dplyr::select(time_span, Lat, Long, elevation, AREA) %>% 
    mutate(metric = THE_metric) %>% 
    crossing(period = c("M2M3", "M3M4", "M2M4")) %>% 
    cbind(pred = round(pred))
  )
}
## Compute turnover
predicted_nCEP_BBS <- 
predicted_nCEP_BBS %>% 
  pivot_wider(id_cols = c(Lat, Long, AREA, elevation, time_span),
              names_from = c(metric,period),
              values_from = pred,
              names_sep = "") %>% 
  group_by(Lat, Long, AREA, elevation, time_span) %>% 
  mutate(jaccardM2M3 = persM2M3/(persM2M3 + colM2M3 + extM2M3),
         jaccardM3M4 = persM3M4/(persM3M4 + colM3M4 + extM3M4),
         jaccardM2M4 = persM2M4/(persM2M4 + colM2M4 + extM2M4)) %>% 
  ungroup()
```

@@

## Number of Colonization, Extinction, Persistence and Turnover

```{r}
pdf("../figures/nCEP_scaling.pdf",
    height = 5, width = 5.5)
for(periods in c("M2M3", "M3M4", "M2M4")){
  print(
    predicted_nCEP_atlas %>% dplyr::select(spatial_scale, contains("M", ignore.case = F)) %>% dplyr::rename(AREA = spatial_scale) %>% 
    rbind(
      predicted_nCEP_BBS %>% dplyr::select(contains("M", ignore.case = F), AREA) %>% 
        mutate(AREA = ifelse(AREA < 0.2, 0.13, 2.51))
    ) %>% 
      pivot_longer(cols = -c("AREA"),
                 names_to = "period",
                 values_to = "value") %>% 
    separate(period, into = c("metric", "period"), sep = c(-4)) %>% 
    filter(metric == "col" | metric == "ext" | metric == "pers") %>% 
  filter(period == periods) %>% 
  ggplot()+
  geom_boxplot(aes(AREA, value,
                   group = interaction(AREA,metric),
                   fill = metric),
                   outlier.shape = NA)+
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")+
  scale_x_log10()+
  scale_y_log10(limits = c(4,210))+
  annotation_logticks(sides = "bl")+
  stat_summary(aes(AREA, value, color = metric), fun.y = median, geom = 'line', linetype = 2, size = 1)+
  ylab("Count")+
  xlab("Spatial grain\n(km²)")+
  labs(fill = "Probability")+
  theme_bw()+
  theme(text = element_text(size = 16),
        legend.position = "none")
  )
}
dev.off()
```

```{r}
pdf("../figures/col_ext_2periods.pdf",
    height = 5, width = 5.5)

for(THE_metric in c("col", "ext")){
print(predicted_nCEP_atlas %>% dplyr::select(spatial_scale, contains("M", ignore.case = F)) %>%
        dplyr::rename(AREA = spatial_scale) %>% 
        rbind(predicted_nCEP_BBS %>% dplyr::select(contains("M", ignore.case = F), AREA) %>% 
                mutate(AREA = ifelse(AREA < 0.2, 0.13, 2.51))) %>% 
        dplyr::select(AREA, contains(THE_metric)) %>% 
        pivot_longer(cols = contains(THE_metric),
                     names_to = "period",
                     values_to = "value") %>% 
        mutate(period = str_remove(period, THE_metric)) %>% 
          mutate(period = case_when(
            period == "M2M4" ~ "1985-2017",
            period == "M2M3" ~ "1985-2003",
            period == "M3M4" ~ "2003-2017")) %>% 
        mutate(period = fct_relevel(period, c("1985-2017", "1985-2003", "2003-2017"))) %>%
        filter(period != "1985-2017") %>%
        ggplot()+
        geom_boxplot(aes(AREA, value, group = interaction(AREA, period), color = period))+
        stat_summary(aes(AREA, value, color = period), fun.y = median, geom = 'line', size = 1, linetype = 2)+
        scale_x_log10()+
        annotation_logticks(sides = "b")+
        xlab("Spatial grain (km²)")+
        ylab(THE_metric)+
        ylim(0,50)+
        labs(color = "Period", shape = "Period", linetype = "Period")+
        theme_bw()+
        theme(text = element_text(size = 16),
              legend.position = "none"))
}
dev.off()
```

```{r}
pdf("../figures/jaccard.pdf",
    height = 5, width = 5.5)

for(period in c("jaccardM2M3","jaccardM3M4")){
print(predicted_nCEP_atlas %>% 
        dplyr::select(spatial_scale, contains("M", ignore.case = F)) %>%
        dplyr::rename(AREA = spatial_scale) %>% 
        rbind(predicted_nCEP_BBS %>% 
        dplyr::select(contains("M", ignore.case = F), AREA) %>% 
        mutate(AREA = ifelse(AREA < 0.2, 0.13, 2.51))) %>% 
        dplyr::select(AREA, period) %>% 
        dplyr::rename(value = period) %>% 
        ggplot(aes(AREA, value, group = AREA))+
        annotation_logticks(sides = "b")+
        geom_boxplot()+
        scale_x_log10()+
        xlab("Spatial grain\n(km²)")+
        ylim(.35, 1)+
        ylab("Jaccard")+
        theme_bw()+
        theme(strip.background = element_blank(),
              strip.text.x = element_blank(),
              text = element_text(size = 16))+
        stat_summary(aes(group = 1), fun.y = median,
                     geom = "line", linetype = "dashed", size = 1, color = "blue"))
}
dev.off()
```

>>>>

```{r, message=FALSE, warning=FALSE}
rm(list = ls())
library(caret)
library(caretEnsemble)
library(dplyr)
library(ranger)
library(ggtext)
library(cowplot)
library(forcats)
library(tidyr)
```

@@

## We perform repeated (n = 3) cross-validation (k = 10) using the caret package
# First we split the dataset into training and validation datasets
set.seed(333)
traindata <- trainControl(method = "repeatedcv", number = 10, repeats = 3)

@@

## And make predictions for each metric
predicted_nCEP_atlas <- data.frame()
## predictions for each metric
for(THE_metric in names(nCEP_models)){
  model <- nCEP_atlas_models[[THE_metric]][["xgboost"]]
  pred <- predict(model, 
                  newdata = nCEP_topredict)
  predicted_nCEP_atlas <- 
  rbind(predicted_nCEP_atlas,
        nCEP_topredict %>% cbind(pred = round(pred)) %>% mutate(metric = THE_metric))
}
## Also, compute jaccard index
## Compute turnover
predicted_nCEP_atlas <- 
predicted_nCEP_atlas %>% 
  pivot_wider(id_cols = c(Lat, Long, AREA, polypoint_ratio, KVADRAT, elevation, spatial_scale, effort_diff, timespan_diff),
              names_from = c(metric,period),
              values_from = pred,
              names_sep = "") %>% 
  group_by(Lat, Long, AREA, polypoint_ratio, KVADRAT, elevation, spatial_scale, effort_diff, timespan_diff) %>% 
  mutate(jaccardM2M3 = persM2M3/(persM2M3 + colM2M3 + extM2M3),
         jaccardM3M4 = persM3M4/(persM3M4 + colM3M4 + extM3M4),
         jaccardM2M4 = persM2M4/(persM2M4 + colM2M4 + extM2M4)) %>% 
  ungroup()
```

### JPSP

```{r}

@@
